---
title: "Getting started with ukbcareer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ukbcareer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This walks through the whole pipeline on the synthetic people that ship with the
package. For real use, replace `ukb_synth()` with your own CSV paths and supply
the three extra fields described in section 2.

## 1. Five minutes to results (Tier 0, no bundle needed)

```{r}
library(ukbcareer)
d <- ukb_synth(n = 200)          # real use: "ukb_cat130.csv"
wl <- ukb_worklife(d, verbose = FALSE)
wl
```

`ukb_career()` chains the tiers together and returns one row per person.
**By default it runs Tier 0 + 1 only**: fast, no dependency on model weights,
no assignment error.

```{r}
out <- ukb_career(d, what = c("core", "exposure"), verbose = FALSE)
dim(out)
out[1:3, .(eid, sex, career_end, career_end_source, work_years,
           exposure_breadth)]
```

Carry `career_end_source` with the data wherever it goes. It separates
**retirees**, for whom `career_end` is a substantive event, from the
**administratively censored**, for whom it is merely the edge of the observation
window. Collapsing the two equates "retired at 53" with "still working when
recruited at 53".

## 2. Which fields you need

Category 130 in full, plus three fields that are **not** in it:

| Field | Where | Why it cannot be skipped |
| --- | --- | --- |
| 22500 date questionnaire completed | **Category 123** | The third term of `career_end` is the per-person questionnaire year. A constant instead adds about 2 years to every ongoing record (5.7% of all person-years) and drops the observed-retirement rate from 56.8% to 27.7%. |
| 31 sex | baseline | The representations are trained separately per sex, and the CAMSIS scale is itself sex-specific. |
| 34 year of birth | baseline | The anchor of the age axis. If the main table also has one, the package **prefers the baseline value** and reports any disagreement. |

```{r eval = FALSE}
wl <- ukb_worklife("ukb_cat130.csv",
                   entry_path     = "ukb_cat123.csv",
                   covariate_path = "ukb_baseline.csv")
```

## 3. Quality control: read it before you use the results

Your sample is not our reference cohort, and **silent drift in the definitions
is this package's most dangerous failure mode**. Every check points at one
specific cause rather than at "data quality" in general:

```{r}
q <- ukb_qc(wl, quiet = TRUE)
q[status != "ok", .(check, value, status)]
```

Given a bundle, each item is also compared against the reference cohort. A
`fail` must be resolved: it means your definitions differ from the paper's and
the results are not comparable. A `warn` is not necessarily an error -- it may
be a real difference in sample selection -- but it has to be explainable.

## 4. Nine-state annual sequences

```{r}
st <- ukb_states(wl)
st
```

The trailing `NA` **is information, not missingness**: it marks that the career
ended at `career_end`. Hand the matrix to TraMineR with `as_seqdef(st)`.

## 5. Plots

Three levels -- person, sample, reference -- all returning ggplot objects.

```{r fig.width = 7, fig.height = 2.6}
plot_career(wl, eid = wl$cohort$eid[4])
```

```{r fig.width = 7, fig.height = 4}
plot_chronogram(wl, by = "sex", which = "absolute")
```

On the absolute-share panel the between-group differences are almost invisible,
because full-time work dominates employed person-years. Given a bundle,
`plot_chronogram()` also draws the **difference** from the reference cohort at
the same age, which is where the information actually is.

Save through `ukb_save()`, the package's single exit point (uniform PDF, font
embedding, and a sane dpi for the rasterised layers):

```{r eval = FALSE}
ukb_save(plot_flows(wl, "Female"), "flows.pdf", width = 9, height = 7)
```

## 6. Tier 1 and Tier 2

```{r eval = FALSE}
b <- ukb_bundle("~/ukbcareer_bundle_v5.0")
ukb_bundle_info(b)          # **read this before using `type`**

out <- ukb_career(wl, bundle = b,
                 what = c("core", "exposure", "movement", "latent", "type"))
```

Three things you must know. Details in `?ukb_typology`, `?ukb_movement` and the
README.

1. **Your types are not the paper's types.** Leiden is transductive, so the
   package re-runs the published recipe on your sample instead of assigning you
   to our 12 (female) / 9 (male) clusters. What is comparable across studies is
   the method and the representation, not the partition.
2. **Check the seed sensitivity.** `ukb_typology()` re-runs with several seeds
   and reports the median pairwise ARI. On a sample of about 1,000 people we
   measured 0.28 to 0.73 -- the data did not change and the partition did.
3. **Mobility geometry is smaller than it looks.** Pairwise distances between
   the 353 occupation codes are highly concentrated (CV 0.10), so
   `path_length`, `detour_ratio` and `straightness` are all the number of
   changes in disguise. The package does not return them.

Per-person perplexity is not returned either. The encoder is bidirectional, so
per-token loss measures interpolation rather than forecasting: the model gains
only 0.7% over a rule that copies the two adjacent years, and per-person values
have an interquartile range of 0.0008. `plot_baselines()` shows this.
