---
title: "Getting started with ukbcareer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ukbcareer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.align = "center", out.width = "100%")
has_plots <- requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("patchwork", quietly = TRUE)
```

This vignette takes you from UK Biobank CSV files to a first analysis. Every
chunk that runs here uses the **synthetic participants** that ship with the
package, so you can follow along without any UK Biobank data. Where your own
files come in, the code is shown but not run.

## 1. Try it on synthetic data

```{r}
library(ukbcareer)

d <- ukb_synth(n = 200)     # 200 synthetic people in the UK Biobank layout
res <- ukb_career(d, verbose = FALSE)
res
```

`res` is a `data.table` with one row per person. The printout shows a few key
columns; `View(res)` shows them all.

To find out what a column means, ask the dictionary:

```{r}
ukb_dictionary("career_end")
ukb_dictionary("yrs_high_asbestos")
```

`ukb_dictionary()` with no arguments lists every variable, and
`vignette("variables")` explains each group in more depth.

## 2. Look at one person

Before analysing thousands of people, it helps to look at a few. `plot_person()`
draws a participant's whole work history on one page:

```{r fig.width = 9, fig.height = 8.5, eval = has_plots}
plot_person(res, eid = attr(d, "showcase_eid"))
```

How to read it, from top to bottom:

- **state**: what the person was doing each year (full-time, part-time,
  unemployed, looking after home and family, education, health problems,
  retired, ...).
- **SOC 4-digit ... 1-digit**: the occupation code of each job, from the most
  detailed to the broadest level of the SOC2000 classification.
- **gap code**: the reason for each break between jobs.
- **dt**: how long the person had been in the same occupation (capped at 3
  years).
- **jobs + flags**: the number of jobs held that year; a triangle marks a
  change of occupation, a square marks overlapping jobs.
- **exposure intensity**: how often each workplace exposure was reported, plus
  hours and shift work. Each row is scaled to its own maximum; `?` marks
  "do not know".
- **Hatched** cells are years not in work, where occupation and exposure are
  undefined. The red dashed line is `career_end`; nothing after it is analysed.

To make a PDF with one page per person, for example to check a random sample
of records, use `ukb_report()`:

```{r eval = FALSE}
set.seed(1)
ukb_report(res, eid = sample(res$eid, 50), file = "check_50_people.pdf")
```

## 3. Your own UK Biobank data

You need three exports from your own application:

| File | Contents |
| --- | --- |
| work history | Category 130 in full (jobs, gaps, hours, shifts, exposures) |
| questionnaire date | field 22500 "When occupational data entered", in **Category 123** |
| baseline | field 31 (sex) and field 34 (year of birth) |

Field 22500 is the one people most often miss because it is not in Category
130. Without it the package falls back to a fixed year, which adds about two
years to every job that was ongoing at the questionnaire and roughly halves the
share of people with an observed retirement.

```{r eval = FALSE}
res <- ukb_career("ukb_cat130.csv",
                  entry_path     = "ukb_cat123.csv",
                  covariate_path = "ukb_baseline.csv")
```

Columns are expected to be named like `Year_job_started__0_3` (field title,
instance, array index). If your export names them differently, pass a mapping
through `ukb_worklife(fields = list(...))` and give the result to
`ukb_career()`.

### Check the data first

`ukb_qc()` runs a set of checks, each aimed at one specific way the input can
go wrong (for example, negative exposure codes that were cleaned away as
missing, or a missing questionnaire date):

```{r}
q <- ukb_qc(res, quiet = TRUE)
q[, .(check, value, status)]
```

Anything marked `fail` must be fixed before the results are usable. A `warn`
is not necessarily an error, but you should be able to explain it.

## 4. A first analysis

Merge the variables with your outcomes and covariates by `eid`:

```{r eval = FALSE}
library(data.table)
outcomes <- fread("my_outcomes.csv")          # eid, outcome, covariates ...
dat <- merge(outcomes, res, by = "eid")
```

A few patterns that come up often (see `vignette("variables")` for more):

```{r}
# How did careers end?
res[, .N, by = career_end_source]

# Working years and shift work by sex
res[, .(median_work_years = median(work_years, na.rm = TRUE),
        any_shift_work = mean(shift_years > 0, na.rm = TRUE)), by = sex]

# Sensitivity analysis: drop poorly documented histories
res_complete <- res[flag_incomplete == 0]
nrow(res_complete)
```

```{r eval = FALSE}
# Exposure as cumulative dose, per 10 years at 'often'
fit <- glm(outcome ~ I(yrs_high_dusty / 10) + work_years + year_birth + sex,
           family = binomial, data = dat)

# Time to retirement: only "retire" is an event, "cap" is censored
library(survival)
fit_ret <- coxph(Surv(career_end, career_end_source == "retire") ~ sex,
                 data = dat)
```

## 5. Whole-sample pictures

```{r fig.width = 7, fig.height = 4, eval = has_plots}
plot_states(res, max_rows = 200)
```

```{r fig.width = 7, fig.height = 4, eval = has_plots}
plot_chronogram(res, by = "sex", which = "absolute")
```

`plot_exposure(res)` summarises the ten exposures by group, and
`plot_career(res, eid)` gives a compact one-line version of the person report.
All plots are ggplot objects; `ukb_save()` writes them to PDF or PNG with
sensible defaults:

```{r eval = FALSE}
ukb_save(plot_states(res), "states.pdf", width = 7, height = 4)
```

## 6. Sequence analysis

The nine-state annual sequences can go straight into TraMineR:

```{r}
st <- ukb_states(res)
st
```

```{r eval = FALSE}
library(TraMineR)
sq <- as_seqdef(st)
seqdplot(sq)
```

Empty cells at the end of a sequence mean the career had already ended
(`career_end`); they are not missing data.

## 7. Occupational status (CAMSIS)

CAMSIS scores measure the social advantage of an occupation. The table is not
distributed with the package; obtain a SOC2000 version with columns
`soc2000`, `mcamsis` and `fcamsis`, then:

```{r eval = FALSE}
res <- ukb_career("ukb_cat130.csv", entry_path = "ukb_cat123.csv",
                  covariate_path = "ukb_baseline.csv",
                  camsis_path = "camsis_soc2000.csv",
                  what = c("core", "exposure", "camsis"))

plot_flows(res, "Female")    # occupational flows over the life course
```

## 8. Mobility and career types (optional model)

Occupational mobility, the career representation and career types use a model
pretrained on the UK Biobank reference cohort. It is available on request to
researchers with their own UK Biobank application (see `BUNDLE-ACCESS.md` on
the package home page).

```{r eval = FALSE}
b <- ukb_bundle("~/ukbcareer_bundle_v5.0")   # verifies every file
ukb_bundle_info(b)                            # read this before using `type`

res <- ukb_career("ukb_cat130.csv", entry_path = "ukb_cat123.csv",
                  covariate_path = "ukb_baseline.csv", bundle = b,
                  what = c("core", "exposure", "movement", "type"))

plot_type_mix(res)          # what each career type looks like
```

Two things to keep in mind with career types:

1. **They are found in your sample.** The clustering is re-run on your data
   with the same settings as the paper, so your type numbers and their meaning
   are your own. Do not match your type 3 against the paper's.
2. **Check how stable they are.** `ukb_typology()` repeats the clustering with
   several random seeds and warns when the results disagree; the numbers are in
   `attr(types, "diagnostics")`. Describe each type before putting it in a
   model, and prefer the continuous variables for primary analyses.
