---
title: "The derived variables: what they are and how to use them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The derived variables: what they are and how to use them}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

`ukb_career()` returns one row per participant. This vignette goes through the
variables group by group: what each one measures, how to put it into a model,
and the mistakes to avoid. The same information, variable by variable, is
available at any time from `ukb_dictionary()`.

```{r}
library(ukbcareer)
d <- ukb_synth(n = 300)
res <- ukb_career(d, verbose = FALSE)
```

## Where the variables come from

UK Biobank participants reported, in an online questionnaire (field 22500 gives
the date), every job they had held (start and end year, occupation, hours,
shift work, ten workplace exposures) and every gap between jobs with its
reason. The package turns this into an **annual record from age 16 to the end
of the career**, and summarises that record per person.

| Group | Needs | Get it with |
| --- | --- | --- |
| Career timing, record quality | your CSVs | `what = "core"` (default) |
| Workplace exposures, working hours and shifts | your CSVs | `what = "exposure"` (default) |
| Occupational status (CAMSIS) | a CAMSIS table | `what = "camsis"` + `camsis_path` |
| Occupational mobility | the model bundle | `what = "movement"` + `bundle` |
| Career type | the model bundle + `torch`, `igraph` | `what = "type"` + `bundle` |
| Career representation | the model bundle + `torch` | `include_latent = TRUE` + `bundle` |

## Career timing

`career_end` is the age at which the working career ends. It is the earliest
of three ages: a definite retirement, the year after the last job, and the age
at the questionnaire. Every other summary (working years, exposure years, ...)
is counted up to `career_end`.

**`career_end_source` says which of the three it was, and it changes the
meaning completely:**

| `career_end_source` | Meaning | In a time-to-event analysis |
| --- | --- | --- |
| `retire` | the person reported retiring | event |
| `cap` | still working at the questionnaire | censored |
| `last_job_plus1` | stopped working, no retirement recorded | ambiguous: try both |

```{r}
res[, .(n = .N, median_end = median(career_end)), by = career_end_source]
```

`retire_age` is the age at the first definite retirement and `observed_retirement`
flags whether there was one. They are not the same as `career_end_source ==
"retire"`: someone can report retirement years after their last job, in which
case `career_end` comes from the last job. `work_years` counts the years with at
least one job and is the natural measure of career length. `max_parallel` is
the largest number of jobs held in any one year.

## Record quality

Not everyone documented their whole working life. `coverage` is the share of
years between 16 and `career_end` that are covered by a job or gap record;
`flag_incomplete` marks people with `coverage < 0.5`.

```{r}
summary(res$coverage)
res[, .N, by = flag_incomplete]
```

A common approach is to run the main analysis on everyone and a sensitivity
analysis without `flag_incomplete == 1`. `first_year` and `last_year` are the
first and last ages with a record (ages, despite the names); a late
`first_year` means the early career was not reported.

## Workplace exposures

For each job, participants said how often their workplace was very noisy,
very cold, very hot, very dusty, full of fumes or other people's cigarette
smoke, or involved asbestos, paints/thinners/glues, pesticides or diesel
exhaust: *rarely/never*, *sometimes* or *often*. For each of the ten agents
there are four summaries:

| Variable | Meaning | Typical use |
| --- | --- | --- |
| `yrs_exp_<agent>` | working years at *sometimes* or *often* | cumulative dose |
| `yrs_high_<agent>` | working years at *often* | cumulative dose at the highest level; the most specific |
| `peak_<agent>` | highest level ever: 0, 1, 2 | ever/never (`peak >= 1`) or ever-high (`peak == 2`) |
| `mean_<agent>` | average level over working years (0-2) | intensity, independent of career length |

```{r}
res[, .(ever_dusty = mean(peak_dusty >= 1, na.rm = TRUE),
        median_yrs_dusty_if_ever = median(yrs_exp_dusty[yrs_exp_dusty > 0],
                                          na.rm = TRUE))]
```

Three overall indices summarise all ten: `exposure_breadth` (how many agents
were ever reported), `high_exposure_years` (agent-years at *often*; can exceed
`work_years` because agents co-occur) and `longest_agent_years` (the longest
heavy exposure to a single agent).

**"Do not know" is kept separate from "never".** When every answer for an
agent was "do not know", `peak_<agent>` and `mean_<agent>` are `NA`, not 0.
In the year counts, "do not know" years count as 0. When you model
`peak_<agent>`, decide explicitly what to do with the `NA`s (exclude, or a
separate category), rather than letting a model silently drop them.

When comparing dose across people with very different career lengths, adjust
for `work_years`, or use `mean_<agent>` instead of the year counts.

## Working hours and shifts

`long_hours_years` counts working years at 48 or more hours a week (the EU
Working Time Directive limit), using the main job of each year. `shift_years`
counts years with any shift work and `night_shift_years` years with night
shifts. `long_hours_frac` and `shift_frac` divide by `work_years`, giving the
share of the working life.

```{r}
res[, .(any_night_shifts = mean(night_shift_years > 0, na.rm = TRUE),
        mean_shift_frac = mean(shift_frac, na.rm = TRUE)), by = sex]
```

## Occupational status (CAMSIS)

CAMSIS scores rank occupations by social advantage (higher = more advantaged),
with separate scales for men and women. With a CAMSIS table
(`camsis_path = ...`) you get:

- `camsis_mean`: lifetime occupational status, the usual choice for a
  life-course socioeconomic position covariate;
- `camsis_first`, `camsis_last`: status at entry and at the end of the career;
- `camsis_change` (`last - first`) and `camsis_slope` (points per year):
  intragenerational mobility;
- `camsis_pattern`: `"rising"`, `"flat"` or `"declining"`, an easy category to
  report; `zero_slope` flags people whose score never changed at all.

`camsis_slope` is `NA` for people with fewer than five scored years or a span
under three years, because a slope cannot be estimated reliably.

## Occupational mobility

These need the model bundle, which provides a map of how similar the 353
occupation codes are to each other.

- `n_moves`: the number of changes of occupation. About a third of people
  never change; this is the backbone mobility measure.
- `mean_jump`: how *far* a typical move goes, independent of how *often*
  (about 19 within the same broad group, about 37 to a different one).
- `max_single_jump`, `net_displacement` (from the first to the last
  occupation), `first_move_age`, `first_move_rel` (years from the first job to
  the first move; less tied to birth year than `first_move_age`) and
  `last_move_age`.

**All distance and timing measures are `NA` when `n_moves == 0`**: the
distance of a move that never happened is undefined, not zero. In a model,
either restrict to people who moved, or enter `n_moves == 0` as its own
indicator and set the distance to a constant for them.

## Career type

With the model bundle, `what = "type"` groups people whose whole careers look
alike. The model turns each career into a numeric summary (the career
representation below), and the same clustering recipe as in the paper is run
on **your** sample, separately by sex. Types are numbered 0, 1, 2, ...

How to use it well:

1. **Describe the types first.** `plot_type_mix()` and `plot_chronogram(res,
   by = "type")` show what each type looks like (states over age, occupations,
   exposures). Name them from what you see.
2. **Your types are yours.** Type numbers and meanings do not carry over from
   the paper or from another study; what is shared is the method.
3. **Check stability.** The grouping is only moderately stable (bootstrap ARI
   0.52 for women and 0.62 for men on the reference cohort). `ukb_typology()`
   warns when changing only the random seed changes the result much.
4. **Treat `type` as a descriptive summary.** Use it as a categorical exposure
   with a clearly defined reference type, and prefer the continuous variables
   above for primary hypotheses.

## Career representation

`include_latent = TRUE` adds `z001` ... `z192`, a 192-number summary of the
whole career. Individual dimensions have no meaning of their own. Use them
together: to find similar careers, to cluster, or as a high-dimensional
adjustment set (for example, after reducing them with principal components).

## The full dictionary

```{r}
dd <- ukb_dictionary()
knitr::kable(dd[, .(variable, group, needs, description)])
```
